For Finitary Induction-Induction, \\ Induction is Enough

### 2020-04-29 paper/paper.tex
stuff

\documentclass[a4paper,UKenglish,cleveref, autoref]{lipics-v2019}
%This is a template for producing LIPIcs articles.
%See lipics-manual.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling cleveref support, use "autoref"
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{hyperref}
\usepackage{todonotes}
\presetkeys{todonotes}{inline}{}
\usepackage{csquotes} %for \llbracket and \rrbracket
\usepackage{stmaryrd}

@@

@@

TODO: list of contributions here.

@@

The current work builds heavily on the work of Kaposi et al.\
\cite{Kaposi:2019:CQI:3302515.3290315} on finitary quotient
inductive-inductive types (QIITs); we reuse both QIIT syntax and
semantics by restricting to IITs, and we reuse the term model
construction of QIITs as well. We also make use of the extesion to
infinitary QIITs \cite{large_inf_qiit} to derive the specification of
the elimination principle for the theory of IIT signatures.

@@

\begin{definition}[Model of extensional type theory (ETT)]
  By a model of ETT we mean a category with families (CwF)
  \cite{Dybjer96internaltype,Hofmann97syntaxand} with an infinite
  predicative hierarchy of universes closed under the following type
  formers: $\Pi$, $\Sigma$, $0$, $1$, $2$, indexed W-types
  \cite{indexedcont} and an identity type with uniqueness of identity
  proofs and equality reflection.
\end{definition}

We will use Agda-like type theoretic syntax to work in the internal
language of models of ETT:
\begin{itemize}
\item Universes are written $\Set_i$. We usually omit level indices in this paper.
\item $\Pi$ types are notated as $(x : A)\ra B$, or as $A \ra B$ when
  non-dependent. We sometimes omit function arguments, by implicitly
  generalising over variables.
\item $\Sigma$-types, notated either as $(x : A)\times B$, or as
  $\sum\limits_{x} B$ when we want to leave the type of the first
  projection implicit. We use $A \times B$ for non-dependent pairs.
\item We will define indexed W-types \cite{indexedcont} by listing
  their constructors and using pattern matching to eliminate from
  them.
\item The equality (identity) type is written $t = u$, it has a
  constructor $\refl : t = t$, and equality reflection, hence we use
  the same $=$ sign for definitional equality. We occasionally
  indicate by $e_1,\dots,e_n \#t$ that $t$ is well-typed thanks to the
  equalities $e_1$,\dots,$e_n$. We also have uniqueness of identity
  proofs (UIP), expressing $(e : t = t)\ra e = \refl$. Note that
  function extensionality, expressing
  $((x : A) \ra f\,x = g\,x) \ra f = g$ is derivable.
\end{itemize}

@@

In this section we describe signatures for IITs. We will say what it
means that a model of ETT supports the theory of IIT signatures. The
theory of IIT signatures is a particular QIIT, and supporting a QIIT
means that we have a notion of algebra and morphism, and there is an
initial algebra with a unique morphism to every other algebra (called
recursor or non-dependent eliminator).

@@

\begin{definition}[Signature algebra]\label{def:algebra}
  In a model of ETT, a signature algebra consists of the following
  four (families of) sets, 17 operations and 18 equalities written in
  the internal language.
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\
  & \Con && : \Set \\
  & \Ty  && : \Con\ra\Set \\
  & \Sub  && : \Con\ra\Con\ra\Set \\
  & \Tm  && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Set \\
  & \id && : \Sub\,\Gamma\,\Gamma \\
  & \blank\circ\blank && : \Sub\,\Theta\,\Delta\ra\Sub\,\Gamma\,\Theta\ra\Sub\,\Gamma\,\Delta \\
  & \ass && : (\sigma \circ \delta) \circ \nu = \sigma \circ (\delta \circ \nu) \\
  & \idl && : \id\circ\sigma = \sigma \\
  & \idr && : \sigma\circ\id = \sigma \\
  & \blank[\blank] && : \Ty\,\Delta\ra\Sub\,\Gamma\,\Delta\ra\Ty\,\Gamma \\
  & \blank[\blank] && : \Tm\,\Delta\,A\ra(\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma]) \\
  & [\id] && : A[\id] = A \\
  & [\circ] && : A[\sigma \circ \delta] = A[\sigma][\delta] \\
  & [\id] && : t[\id] = t \\
  & [\circ] && : t[\sigma \circ \delta] = t[\sigma][\delta] \\
  & \boldsymbol{\cdot} && : \Con \\
  & \epsilon && : \Sub\,\Gamma\,\boldsymbol{\cdot} \\
  & {\boldsymbol{\cdot}\eta} && : (\sigma : \Sub\,\Gamma\,\boldsymbol{\cdot}) \ra \sigma = \epsilon \\
  & \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
  & \blank,\blank && : (\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma])\ra\Sub\,\Gamma\,(\Delta\ext A) \\
  & \pi_1 && : \Sub\,\Gamma\,(\Delta\ext A)\ra \Sub\,\Gamma\,\Delta \\
  & \pi_2 && : (\sigma : \Sub\,\Gamma\,(\Delta\ext A))\ra \Tm\,\Gamma\,(A[\pi_1 \sigma]) \\
  & {\pi_1\beta} && : \pi_1(\sigma, t) = \sigma \\
  & {\pi_2\beta} && : \pi_2(\sigma, t) = t \\
  & {\pi\eta} && : (\pi_1\,\sigma, \pi_2\,\sigma) = \sigma \\
  & {,\circ} && : (\sigma, t)\circ\delta = (\sigma\circ\delta, t[\delta]) \\
  & \rlap{(2) Universe} \\
  & \U && : \Ty\,\Gamma \\
  & \El && : \Tm\,\Gamma\,\U\ra\Ty\,\Gamma \\
  & {\U[]} && : \U[\sigma] = \U \\
  & {\El[]} && : (\El\,a)[\sigma] = \El\,(a[\sigma]) \\
  & \rlap{(3) Inductive parameters} \\
  & \Pi && : (a:\Tm\,\Gamma\,\U)\ra\Ty\,(\Gamma\ext\El\,a)\ra\Ty\,\Gamma \\
  & \blank\oldapp\blank && : \Tm\,\Gamma\,(\Pi\,a\,B)\ra (u : \Tm\,\Gamma\,(\El\,a))
  \ra \Tm\,\Gamma\,(\El\,(B[\id,\,u])) \\
  & {\Pi[]} && : (\Pi\,a\,B)[\sigma] = \Pi\,(a[\sigma])\,(B[\sigma\circ\p,\q]) \\
  & {\oldapp[]} && : (t\oldapp\alpha)[\sigma] = (t[\sigma])\mathop{\oldapp}(\alpha[\sigma]) \\
  & \rlap{(4) External parameters} \\
  & \Pim && : (T:\Set)\ra(T\ra\Ty\,\Gamma)\ra\Ty\,\Gamma \\
  & \blank\appm \blank && : \Tm\,\Gamma\,(\Pim\,T\,B)\ra(\alpha:T) \ra\Tm\,\Gamma\,(B\,\alpha) \\
  & {\Pim[]} && : (\Pim\,T\,B)[\sigma] = \Pim\,T\,(\lambda \alpha.(B\,\alpha)[\sigma]) \\
  & {\appm[]} && : (t\appm\alpha)[\sigma] = (t[\sigma])\mathop{\appm}\alpha
\end{alignat*}
\end{definition}

The definition of morphism is determined by the notion of algebra, it
can be computed following \cite{large_inf_qiit}, but we include it
here for completeness.
\begin{definition}[Signature morphism]\label{def:morphism}
  A morphism between signature algebras $M$ and $N$ consists of four
  functions and 17 equalities expressing that the functions preserve
  the operations of the two algebras. We use the same naming as in
  \Cref{def:algebra} and use superscripts to denote which algebra is
  meant.
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\
  & \Con && : \Con^M && \ra\Con^N \\
  & \Ty  && : \Ty^M\,\Gamma && \ra\Ty^N\,(\Con\,\Gamma) \\
  & \Sub  && : \Sub^M\,\Gamma\,\Delta && \ra\Sub^N(\Con\,\Gamma)\,(\Con\,\Delta) \\
  & \Tm  && : \Tm^M\,\Gamma\,A && \ra\Tm^N\,(\Con\,\Gamma)\,(\Ty\,A) \\
  & \id && : \Sub\,\id^M && = \id^N \\
  & {\circ} && : \sigma\circ^M\delta && = \Sub\,\sigma\circ^N\Sub\,\delta \\
  & [] && : A[\sigma]^M && = \Ty\,A[\Sub\,\sigma]^N \\
  & [] && : t[\sigma]^M && = \Tm\,t[\Sub\,\sigma]^N \\
  & \boldsymbol{\cdot} && : \Con\,\boldsymbol{\cdot}^M && = \boldsymbol{\cdot}^N \\
  & \epsilon && : \Sub\,\epsilon^M && = \epsilon^N \\
  & \ext && : \Con\,(\Gamma\ext^M A) && = \Con\,\Gamma\ext^N\Ty\,A \\
  & , && : \Sub\,(\sigma,^M t) && = \Sub\,\sigma ,^N\Tm\,t \\
  & \pi_1 && : \Sub\,(\pi_1^M\,\sigma) && = \pi_1^N\,(\Sub\,\sigma) \\
  & \pi_2 && : \Tm\,(\pi_2^M\,\sigma) && = \pi_2^N\,(\Sub\,\sigma) \\
  & \rlap{(2) Universe} \\
  & \U && : \Ty\,\U^M && = \U^N \\
  & \El && : \Ty\,(\El^M\,a) && = \El^N\,(\Tm\,a) \\
  & \rlap{(3) Inductive parameters} \\
  & \Pi && : \Ty\,(\Pi^M\,a\,B) && = \Pi^N\,(\Tm\,a)\,(\Ty\,B) \\
  & \oldapp && : \Tm\,(t\mathbin{{\oldapp}^M} u) && = \Tm\,t\mathbin{{\oldapp}^N}\,\Tm\,u \\
  & \rlap{(4) External parameters} \\
  & \Pim && : \Ty\,(\Pim^M\,T\,B) && = \Pim^N\,T\,(\lambda\alpha.\Ty\,(B\,\alpha)) \\
  & \appm && : \Tm\,(t\mathbin{{\appm}^M}\alpha) && = \Tm\,t\mathbin{{\appm}^N}\alpha
\end{alignat*}
\end{definition}
\begin{definition}\label{def:theoryofsignatures}
  A model of ETT supports the theory of IIT signatures if there is a
  signature algebra equipped with a unique morphism into any
  algebra. We call this algebra the syntax or initial algebra, the
  morphism is called recursor.
\end{definition}
\begin{definition}[Abbreviations]\label{def:abbrevs}
  For any signature algebra, we use
  $\wk : \Sub\,(\Gamma\ext A)\ra \Gamma$ to mean $\pi_1\,\id$. We
  recover de Bruijn indices by having $0= \pi_2 \,\id$ and
  $1+n = n[\wk]$. $\Pi\,a\,(B[\wk])$ is abbreviated by $a\Ra B$,
  $\Pim\,T\,(\lambda\_.B)$ by $T\Ram B$.
\end{definition}

In a model supporting the theory of IIT signatures, we can write down
signatures as elements of $\Con^\S$ (the $\Con$ component in the
syntax). For example, natural numbers are given by the following
signature. For readability, on the right we list the version of the
signature using variable names instead of CwF-combinators.
\[
\boldsymbol{\cdot}\,\ext\,\U\,\ext\,\El\,0\,\ext\,1 \Ra \El\,1 \hspace{5.9em}
\boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \Ra \El\,N
\]
Natural numbers are given by a context of length 3, starting with the
empty context $\boldsymbol{\cdot}$ and declaring a sort by $\U$,
saying that we have an operator producing an element of $\El\,0$ where
$0$ is a de Bruijn index referring to the sort $\U$. Finally we have
an operator which takes as input an element of the sort $1$ and
produces an element of the same sort. Note the assymetry of the
function type $\Ra$: on the left there needs to be an element of $\U$,
while on the right any type can appear (including another function
type). This ensures strict positivity of the operators.

Lists with elements of a given $T : \Set$ type are given by the
following signature. Here we use the function space with external domain $\Ram$ to include
an $T$ in the signature.
\begin{alignat*}{10}
 & \boldsymbol{\cdot}\,\ext\,\U\,\ext\,\El\,0\,\ext\,T\Ram 1\Ra \El\,1 \hspace{3.56em} \boldsymbol{\cdot}\,\ext\,L:\U\,\ext\,nil : \El\,L\,\ext\,cons : T\Ram L\Ra \El\,L
\end{alignat*}
The $\Con$--$\Ty$ example from \Cref{sec:intro} is specified as
follows.
\begin{alignat*}{5}
  & \boldsymbol{\cdot} \,\ext                                                 && \boldsymbol{\cdot} \,\ext                                                                \\
  & \U\,\ext                                                                  && Con && : \U\,\ext                                                                        \\
  & 0\Ra\U\,\ext                                                              && Ty && : Con\Ra\U\,\ext                                                                   \\
  & \El\,1\,\ext                                                              && empty && : \El\,Con\,\ext                                                                \\
  & \Pi\,2\,(2\oldapp 0 \Ra \El\,3)\ext                                       && ext && : \Pi\,(\Gamma : Con)\,(Ty\oldapp\Gamma \Ra \El\,Con)\ext                         \\
  & \Pi\,3\,(\El\,(3\oldapp 0))\,\ext                                         && U && : \Pi\,(\Gamma:Con)\,(\El\,(Ty\oldapp\Gamma))\,\ext                                 \\
  & \Pi\,4\,(\El\,(4\oldapp(2\oldapp 0\oldapp (1\oldapp 0)))) \hspace{4.79em} && El && : \Pi\,(\Gamma : Con)\,(\El\,(Ty\oldapp(ext\oldapp\Gamma\oldapp (U\oldapp\Gamma))))
\end{alignat*}

%% \todo{Write this section}

%% Show that the theory of IIT signatures as described in Section
%% \ref{sec:theory_of_signatures} exists, that is, it can be defined only
%% using indexed inductive types and a $\Prop$ sort.

%% Ambroise has done this in Agda with funext and UIP using an
%% extrinsic--intrinsic translation.

%% Ambroise formalisation: \url{https://github.com/amblafont/UniversalII/tree/cwf-syntax/Cwf}.

%% In this formalization, there is a discrepancy for the typing rule of application
%% compared to the one of Section~\ref{sec:theory_of_signatures}: it is not
%% \[
%%    \app  : \Tm\,\Gamma\,(\Pi\,a\,B)\ra\Tm\,(\Gamma\ext \El\,a)\,B
%% \]
%% but
%% \[
%%    \app  : \Tm\,\Gamma\,(\Pi\,a\,B)\ra(u : \Tm\,\Gamma\,(\El\,a))\ra\,\Tm\,\Gamma\,B[<u>].
%% \]
%% Actually, I started with the first version, but at some point I
%% changed for the second one because it made some stuff easier (although
%% I don't remember exactly what it was).

%% I think this should be doable with only UIP (or a $\Prop$-valued
%% equality type?) and not funext. This would be important for a
%% computational justification of IITs.

%% Unfortunately, it seems that funext is needed (see the use of funext in
%% Syntax.agda) for the metatheoretic/infinitary
%% parameters part (4-5), for example in the case of the metatheoretic
%% $ \Pim  : (T:\Set)\ra(T\ra\Ty\,\Gamma)\ra\Ty\,\Gamma $. This needs to be
%% investigated more carefully.

>>>>

\documentclass[a4paper,UKenglish,cleveref, autoref]{lipics-v2019}
%This is a template for producing LIPIcs articles.
%See lipics-manual.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling cleveref support, use "autoref"
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{hyperref}
\usepackage{todonotes}
\presetkeys{todonotes}{inline}{}
\usepackage{csquotes} %for \llbracket and \rrbracket
\usepackage{stmaryrd}
% \usepackage{pdflscape} % if we want \begin{landscape} ... \end{landscape}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

TODO: list of contributions here. Notion of signature free from encoding.

@@

The current work builds heavily on the work of Kaposi et al.\
\cite{Kaposi:2019:CQI:3302515.3290315} on finitary quotient
inductive-inductive types (QIITs); we reuse both QIIT syntax and
semantics by restricting to IITs, and we reuse the term model
construction of QIITs as well. We also make use of the extesion to
infinitary QIITs \cite{large_inf_qiit} to derive the specification of
the elimination principle for the theory of IIT signatures. TODO: also
for specifying signatures.

@@

\begin{definition}[Model of extensional type theory (ETT)]
  By a model of ETT we mean a category with families (CwF)
  \cite{Dybjer96internaltype,Hofmann97syntaxand} with an infinite
  predicative hierarchy of universes closed under the following type
  formers: $\Pi$, $\Sigma$, $\bot$, $\top$, $\Bool$, indexed W-types
  \cite{indexedcont} and an identity type with uniqueness of identity
  proofs and equality reflection. TODO: remove indexed W-types, bot
  and bool.
\end{definition}

We will use Agda-like type theoretic syntax to work in the internal
language of models of ETT:
\begin{itemize}
\item Universes are written $\Set_i$. We usually omit level indices in this paper.
\item $\Pi$ types are notated as $(x : A)\ra B$, or as $A \ra B$ when
  non-dependent. We sometimes omit function arguments, by implicitly
  generalising over variables.
\item $\Sigma$-types, notated either as $(x : A)\times B$, or as
  $\sum\limits_{x} B$ when we want to leave the type of the first
  projection implicit. Projections are either named or given by
  $\proj_1$ and $\proj_2$. We use $A \times B$ for non-dependent
  pairs.
\item We will define indexed W-types \cite{indexedcont} by listing
  their constructors and using pattern matching to eliminate from
  them.
\item The equality (identity) type is written $t = u$, it has a
  constructor $\refl : t = t$, and equality reflection, hence we use
  the same $=$ sign for definitional equality. We occasionally
  indicate by $e_1,\dots,e_n \#t$ that $t$ is well-typed thanks to the
  equalities $e_1$,\dots,$e_n$. We also have uniqueness of identity
  proofs (UIP), expressing $(e : t = t)\ra e = \refl$. Note that
  function extensionality, expressing
  $((x : A) \ra f\,x = g\,x) \ra f = g$ is derivable.
\end{itemize}

@@

TODO: rewrite.

In this section we show how to describe signatures for IITs using a
QIIT, the theory of IIT signatures. We will say what it means for a
model of ETT to support the theory of IIT signatures. First we need to
say what a \emph{theory of IIT signature algebra} is and what
morphisms between such algebras are. Then supporting the theory of IIT
signatures means that there exists an algebra with a unique morphism
to every other algebra.

@@

\begin{definition}[Signature algebra, $\SignAlg$]\label{def:algebra}
  In a model of ETT, a signature algebra is an iterated $\Sigma$ type
  consisting of the following components: four (families of) sets, 17
  operations and 18 equalities.
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\
  & \Con && : \Set \\
  & \Ty  && : \Con\ra\Set \\
  & \Sub  && : \Con\ra\Con\ra\Set \\
  & \Tm  && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Set \\
  & \id && : \Sub\,\Gamma\,\Gamma \\
  & \blank\circ\blank && : \Sub\,\Theta\,\Delta\ra\Sub\,\Gamma\,\Theta\ra\Sub\,\Gamma\,\Delta \\
  & \ass && : (\sigma \circ \delta) \circ \nu = \sigma \circ (\delta \circ \nu) \\
  & \idl && : \id\circ\sigma = \sigma \\
  & \idr && : \sigma\circ\id = \sigma \\
  & \blank[\blank] && : \Ty\,\Delta\ra\Sub\,\Gamma\,\Delta\ra\Ty\,\Gamma \\
  & \blank[\blank] && : \Tm\,\Delta\,A\ra(\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma]) \\
  & [\id] && : A[\id] = A \\
  & [\circ] && : A[\sigma \circ \delta] = A[\sigma][\delta] \\
  & [\id] && : t[\id] = t \\
  & [\circ] && : t[\sigma \circ \delta] = t[\sigma][\delta] \\
  & \boldsymbol{\cdot} && : \Con \\
  & \epsilon && : \Sub\,\Gamma\,\boldsymbol{\cdot} \\
  & {\boldsymbol{\cdot}\eta} && : (\sigma : \Sub\,\Gamma\,\boldsymbol{\cdot}) \ra \sigma = \epsilon \\
  & \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
  & \blank,\blank && : (\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma])\ra\Sub\,\Gamma\,(\Delta\ext A) \\
  & \pi_1 && : \Sub\,\Gamma\,(\Delta\ext A)\ra \Sub\,\Gamma\,\Delta \\
  & \pi_2 && : (\sigma : \Sub\,\Gamma\,(\Delta\ext A))\ra \Tm\,\Gamma\,(A[\pi_1 \sigma]) \\
  & {\pi_1\beta} && : \pi_1(\sigma, t) = \sigma \\
  & {\pi_2\beta} && : \pi_2(\sigma, t) = t \\
  & {\pi\eta} && : (\pi_1\,\sigma, \pi_2\,\sigma) = \sigma \\
  & {,\circ} && : (\sigma, t)\circ\delta = (\sigma\circ\delta, t[\delta]) \\
  & \rlap{(2) Universe} \\
  & \U && : \Ty\,\Gamma \\
  & \El && : \Tm\,\Gamma\,\U\ra\Ty\,\Gamma \\
  & {\U[]} && : \U[\sigma] = \U \\
  & {\El[]} && : (\El\,a)[\sigma] = \El\,(a[\sigma]) \\
  & \rlap{(3) Inductive parameters} \\
  & \Pi && : (a:\Tm\,\Gamma\,\U)\ra\Ty\,(\Gamma\ext\El\,a)\ra\Ty\,\Gamma \\
  & \blank\oldapp\blank && : \Tm\,\Gamma\,(\Pi\,a\,B)\ra (u : \Tm\,\Gamma\,(\El\,a))
  \ra \Tm\,\Gamma\,(\El\,(B[\id,\,u])) \\
  & {\Pi[]} && : (\Pi\,a\,B)[\sigma] = \Pi\,(a[\sigma])\,(B[\sigma\circ\p,\q]) \\
  & {\oldapp[]} && : (t\oldapp\alpha)[\sigma] = (t[\sigma])\mathop{\oldapp}(\alpha[\sigma]) \\
  & \rlap{(4) External parameters} \\
  & \Pim && : (T:\Set)\ra(T\ra\Ty\,\Gamma)\ra\Ty\,\Gamma \\
  & \blank\appm \blank && : \Tm\,\Gamma\,(\Pim\,T\,B)\ra(\alpha:T) \ra\Tm\,\Gamma\,(B\,\alpha) \\
  & {\Pim[]} && : (\Pim\,T\,B)[\sigma] = \Pim\,T\,(\lambda \alpha.(B\,\alpha)[\sigma]) \\
  & {\appm[]} && : (t\appm\alpha)[\sigma] = (t[\sigma])\mathop{\appm}\alpha
\end{alignat*}
Given an $M:\SignAlg$, we denote its components as $\Con^M$, $\Ty^M$,
$\Sub^M$, $\Tm^M$, $\id^M$, and so on.
\end{definition}

\begin{definition}[Abbreviations]\label{def:abbrevs}
  For any signature algebra, we use
  $\wk : \Sub\,(\Gamma\ext A)\,\Gamma$ to mean $\pi_1\,\id$. We
  recover de Bruijn indices by setting $0:= \pi_2 \,\id$ and
  $1+n := n[\wk]$. $\Pi\,a\,(B[\wk])$ is abbreviated by $a\Ra B$,
  $\Pim\,T\,(\lambda\_.B)$ by $T\Ram B$.
\end{definition}

\begin{definition}[IIT signature]
  An IIT signature is a context in an arbitrary signature algebra,
  that is, a function $(M:\SignAlg)\ra\Con^M$.
\end{definition}
Note that the notion of IIT signature is relative to a model of ETT:
it is expressed as a term (of a function type) in the model. This is
necessary because of the function space $\Pim$ which has as domain an
arbitrary type in the model. We make use of $\Pim$ in signatures with
external parameters, like the type of the elements in lists.

\begin{example}[Example signatures]\label{ex:signatures}
  The signature for natural numbers is given by the function
  \[
    \lambda M.(\boldsymbol{\cdot}^M\,\ext^M\,\U^M\,\ext^M\,\El^M\,0^M\,\ext^M\,1^M \Ra^M \El^M\,1^M).
  \]
  For readability we write the same signature without the $\lambda$
  and the superscripts below on the left and we also write an informal
  version with variable names on the right for comparison.
  \[
  \boldsymbol{\cdot}\,\ext\,\U\,\ext\,\El\,0\,\ext\,1 \Ra \El\,1 \hspace{5.9em}
  \boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \Ra \El\,N
  \]
  This signature is given by a context of length three. We start with
  the empty context $\boldsymbol{\cdot}$, then we declare a sort by
  $\U$, then we declare an operator producing an element of $\El\,0$
  where $0$ is a de Bruijn index referring to the sort $\U$. Finally
  we declare an operator which takes as input an element of the sort
  $1$ and produces an element of the same sort. Note the assymetry of
  the function type $\Ra$: on the left there needs to be an element of
  $\U$, while on the right any type can appear (including another
  function type). This ensures strict positivity of the operators.

  Lists with elements of a given $T : \Set$ type are given by the
  following signature. Here we use the function space with external
  domain $\Ram$ to include a $T$ in the signature. Again, we omit the
  $\lambda$ and the superscripts for readability.
  \begin{alignat*}{10}
    & \boldsymbol{\cdot}\,\ext\,\U\,\ext\,\El\,0\,\ext\,T\Ram 1\Ra \El\,1 \hspace{3.56em} \boldsymbol{\cdot}\,\ext\,L:\U\,\ext\,nil : \El\,L\,\ext\,cons : T\Ram L\Ra \El\,L
  \end{alignat*}

  The $\Con$--$\Ty$ example from \Cref{sec:intro} is specified as
  follows.
  \begin{alignat*}{5}
    & \boldsymbol{\cdot} \,\ext                                                 && \boldsymbol{\cdot} \,\ext                                                                \\
    & \U\,\ext                                                                  && Con && : \U\,\ext                                                                        \\
    & 0\Ra\U\,\ext                                                              && Ty && : Con\Ra\U\,\ext                                                                   \\
    & \El\,1\,\ext                                                              && empty && : \El\,Con\,\ext                                                                \\
    & \Pi\,2\,(2\oldapp 0 \Ra \El\,3)\ext                                       && ext && : \Pi\,(\Gamma : Con)\,(Ty\oldapp\Gamma \Ra \El\,Con)\ext                         \\
    & \Pi\,3\,(\El\,(3\oldapp 0))\,\ext                                         && U && : \Pi\,(\Gamma:Con)\,(\El\,(Ty\oldapp\Gamma))\,\ext                                 \\
    & \Pi\,4\,(\El\,(4\oldapp(2\oldapp 0\oldapp (1\oldapp 0)))) \hspace{4.79em} && El && : \Pi\,(\Gamma : Con)\,(\El\,(Ty\oldapp(ext\oldapp\Gamma\oldapp (U\oldapp\Gamma))))
  \end{alignat*}
\end{example}

Given a model of ETT and an IIT signature in it, we would like to say
what it means that the model supports the given IIT. For this we
define the signature algebra $\ADS$ which will provide notions of
algebras, displayed algebras and sections for each signature. This is
the same as the $\blank^\A$, $\blank^\D$ and $\blank^\S$ operations in
\cite{Kaposi:2019:CQI:3302515.3290315}. Before defining $\ADS$, we
illustrate its usage by an example.

\begin{example}[Algebras, displayed algebras and sections for natural
  numbers]
  For the signature of natural numbers as given in
  \Cref{ex:signatures}, algebras are given by the $\Sigma$-type
  $(N:\Set)\times N\times(N\ra N)$. A displayed algebra over $(N,z,s)$
  is given by the $\Sigma$-type
  \[
    (N^D:N\ra\Set)\times N^D\,z \times((n:N)\ra N^D\,n\ra N^D\,(s\,n)),
  \]
  and a section of a displayed algebra $(N^D,z^D,s^D)$ over $(N,z,s)$
  is given by the $\Sigma$-type
  \[
    (N^S:(n:N)\ra N^D\,n)\times (N^S\,z = z^D) \times((n:N)\ra N^S\,(s\,n) = s^D\,n\,(N^S\,n)).
  \]
  Displayed algebras over the initial algebra are called motives and
  methods of the eliminator, while a section of a displayed algebra
  over the initial algebra is the eliminator together with its
  computation rules.
\end{example}

\begin{definition}[The signature algebra $\ADS$]\label{def:ads}
  We define an element of $\SignAlg$ by listing all its components
  $\Con$, $\Ty$, $\Sub$, and so on, one per line. Each such components
  has three parts denoted by $^\A$, $^\D$ and $^\S$. The equality
  components of $\SignAlg$ are omitted as they are all reflexivity.
  \begin{alignat*}{10}
    & (\Gamma^\A:\Set) && \hspace{-0.3em}\times && (\Gamma^\D:\Gamma^\A\ra\Set) && \hspace{-0.3em}\times && (\Gamma^\S:(\gamma:\Gamma^\A)\ra\Gamma^\D\,\gamma\ra\Set) \\
    & (A^\A:\Gamma^\A\ra\Set) && \hspace{-0.3em}\times && (A^\D:\Gamma^\D\,\gamma\ra A^\A\,\gamma\ra\Set) && \hspace{-0.3em}\times && (A^\S:\Gamma^\S\,\gamma\,\gamma^D\ra(\alpha:A^\A\,\gamma)\ra \\
    & && && && && \hspace{1em} A^\D\,\gamma^D\,\alpha\ra\Set) \\
    & (\sigma^\A:\Gamma^\A\ra\Delta^\A) && \hspace{-0.3em}\times && (\sigma^\D:\Gamma^\D\,\gamma\ra\Delta^\D\,(\sigma^\A\,\gamma)) && \hspace{-0.3em}\times && (\sigma^\S:\Gamma^\S\,\gamma\,\gamma^D\ra \\
    & && && && && \hspace{1em} \Delta^\S\,(\sigma^\A\,\gamma)\,(\sigma^\D\,\gamma^D)) \\
    & (t^\A:(\gamma:\Gamma^\A)\ra A^\A\,\gamma) && \hspace{-0.3em}\times && (t^\D:(\gamma^D:\Gamma^\D\,\gamma)\ra && \hspace{-0.3em}\times && (t^\S:(\gamma^S:\Gamma^\S\,\gamma\,\gamma^D)\ra \\
    & && && \hspace{1em} A^\D\,\gamma^D\,(t^\A\,\gamma)) && && \hspace{1em}  A^\S\,(t^\A\,\gamma)\,(t^\D\,\gamma^D)) \\
    & \id^\A\,\gamma := \gamma && && \id^\D\,\gamma^D:=\gamma^D && && \id^\S\,\gamma^S := \gamma^S \\
    & (\sigma\circ\delta)^\A\,\gamma := \sigma^\A\,(\delta^\A\,\gamma) && && (\sigma\circ\delta)^\D\,\gamma^D := \sigma^\D\,(\delta^\D\,\gamma^D) && && (\sigma\circ\delta)^\S\,\gamma^S := \sigma^\S\,(\delta^\S\,\gamma^S) \\
    & (A[\sigma])^\A\,\gamma := A^\A\,(\sigma^\A\,\gamma) && && (A[\sigma])^\D\,\gamma^D := A^\D\,(\sigma^\D\,\gamma^D) && && (A[\sigma])^\S\,\gamma^S := A^\S\,(\sigma^\S\,\gamma^S) \\
    & (t[\sigma])^\A\,\gamma := t^\A\,(\sigma^\A\,\gamma) && && (t[\sigma])^\D\,\gamma^D := t^\D\,(\sigma^\D\,\gamma^D) && && (t[\sigma])^\S\,\gamma^S := t^\S\,(\sigma^\S\,\gamma^S) \\
    & \cdot^\A := \top && && \cdot^\D\,\_ := \top && && \cdot^\S\,\_\,\_ := \top \\
    & \epsilon^\A\,\_ := \tt && && \epsilon^\D\,\_ := \tt && && \epsilon^\S\,\_ := \tt \\
    & (\Gamma\rhd A)^\A := && && (\Gamma\rhd A)^\D\,(\gamma,\alpha) :=  && && (\Gamma\rhd A)^\S\,(\gamma,\alpha)\,(\gamma^D,\alpha^D) := \\
    & \hspace{1em} (\gamma:\Gamma^\A)\times A^\A\,\gamma && && \hspace{1em} (\gamma^D:\Gamma^\D\,\gamma)\times A^\D\,\gamma^D\,\alpha && && \hspace{1em} (\gamma^S:\Gamma^\S\,\gamma\,\gamma^D)\times A^\S\,\gamma^S\,\alpha\,\alpha^D \\
    & (\sigma,t)^\A\,\gamma := (\sigma^\A\,\gamma,t^\A\,\gamma) && && (\sigma,t)^\D\,\gamma^D := (\sigma^\D\,\gamma^D,t^\D\,\gamma^D) && && (\sigma,t)^\S\,\gamma^S := (\sigma^\S\,\gamma^S,t^\S\,\gamma^S) \\
    & (\pi_1\,\sigma)^\A\,\gamma := \proj_1\,(\sigma^\A\,\gamma) && && (\pi_1\,\sigma)^\D\,\gamma^D := \proj_1\,(\sigma^\D\,\gamma^D) && && (\pi_1\,\sigma)^\S\,\gamma^S := \proj_1\,(\sigma^\S\,\gamma^S) \\
    & (\pi_2\,\sigma)^\A\,\gamma := \proj_2\,(\sigma^\A\,\gamma) && && (\pi_2\,\sigma)^\D\,\gamma^D := \proj_2\,(\sigma^\D\,\gamma^D) && && (\pi_2\,\sigma)^\S\,\gamma^S := \proj_2\,(\sigma^\S\,\gamma^S) \\
    & \U^\A\,\gamma := \Set && && \U^\D\,\gamma^D\,T := T\ra\Set && && \U^\S\,\gamma^S\,T\,T^D := (\alpha:T)\ra T^D\,\alpha \\
